---
title: "plots"
author: "Mikey Elmers"
date: "9/21/2022"
output: html_document
---

```{r, eval=FALSE, echo=FALSE}
rmarkdown::render(here::here('scripts/03_plots.Rmd'),
                  output_dir = here::here('output/'))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(ggplot2)
library(dplyr)
```

```{r, , echo=FALSE}
df_yale_pints <- read.csv(here::here("data/final/yale_pints.csv")) 
```

## Density plot for each PINT
The following density plots for duration are aggregated for all speakers. Interestingly, I expected to find multiple peaks for the different PINTs but they all have a mostly unimodal distribution.

### Silence
```{r}
df_yale_pints %>% 
  filter(segment == "sil") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```

### Inhalation
```{r}
df_yale_pints %>% 
  filter(segment == "in") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```

### Exhalation
```{r}
df_yale_pints %>% 
  filter(segment == "ex") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```

### Uh
```{r}
df_yale_pints %>% 
  filter(segment == "uh") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```

### Um
```{r}
df_yale_pints %>% 
  filter(segment == "um") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```

### Tongue click
```{r}
df_yale_pints %>% 
  filter(segment == "cl") %>% 
  ggplot(aes(x = dur)) +
  geom_density()
```



## PINTs count
This plot indicates the # of each individual PINT aggregated for all speakers.
```{r}
# aggregate count per pints type
ggplot(df_yale_pints, aes(x = segment)) +
  geom_bar()
```

## Individual PINTs proprotions
This plot shows the proportion for each PINT aggregated for all speakers.
```{r, echo=FALSE}
# calculate total time of all files
total_time = sum(unique(df_yale_pints$filedur))

# other: total time, proportion of total time
other_summary <- df_yale_pints %>% 
  filter(segment == "other") %>% 
  summarize(total_oth = sum(dur), prop_oth = total_oth / total_time * 100)

# silence: total time, proportion of total time
sil_summary <- df_yale_pints %>% 
  filter(segment == "sil") %>%
  summarize(total_sil = sum(dur), prop_sil = total_sil / total_time * 100)

# inhalation: total time, proportion of total time
in_summary <- df_yale_pints %>% 
  filter(segment == "in") %>%
  summarize(total_in = sum(dur), prop_in = total_in / total_time * 100)

# exhalation: total time, proportion of total time
ex_summary <- df_yale_pints %>% 
  filter(segment == "ex") %>% 
  summarize(total_ex = sum(dur), prop_ex = total_ex / total_time * 100)

# uh: total time, proportion of total time
uh_summary <- df_yale_pints %>% 
  filter(segment == "uh") %>% 
  summarize(total_uh = sum(dur), prop_uh = total_uh / total_time * 100)

# um: total time, proportion of total time
um_summary <- df_yale_pints %>% 
  filter(segment == "um") %>% 
  summarize(total_um = sum(dur), prop_um = total_um / total_time * 100)

# click: total time, proportion of total time
cl_summary <- df_yale_pints %>% 
  filter(segment == "cl") %>% 
  summarize(total_cl = sum(dur), prop_cl = total_cl / total_time * 100)
```

```{r, echo=FALSE}
df_yale_prop <- data.frame(segment = c("sil", "in", "ex", "uh", "um", "cl", "other"),
                           prop_total = c(sil_summary$prop_sil, in_summary$prop_in, ex_summary$prop_ex, uh_summary$prop_uh,
                                          um_summary$prop_um, cl_summary$prop_cl, other_summary$prop_oth))

# factor to re-organize order of segments on x-axis
df_yale_prop$segment <- factor(df_yale_prop$segment,
                      levels = c("sil", "in", "uh", "ex", "um", "other", "cl"))
```

```{r}
ggplot(df_yale_prop, aes(x = segment, y = prop_total)) +
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(prop_total, 1)), position = position_dodge(width = 0.9), vjust = -0.5) +
  ylim(0, 15) + 
  labs(x = "PINTs", y = "% of Total Duration") 
```

## Speaker comparison between lectures
This barplot shows the different PINTs usage for each speaker over the three sessions (indicated in color). Speakers tend to be consistent with their PINTs usage across the three sessions. 

Different speakers prefer different PINTs. For example, speaker 1 uses many more "uh" compared to speaker 4. 
```{r, echo=FALSE}
# fact label names for files
sp.lab <- c("Speaker 1", "Speaker 2", "Speaker 3", "Speaker 4", "Speaker 5")
names(sp.lab) <- c(1, 2, 3, 4, 5)
```

```{r, warning=FALSE}
ggplot(df_yale_pints, aes(x = segment, fill = session)) + 
  geom_bar(position = position_dodge()) +
  scale_x_discrete(limits = c("sil", "in", "ex", "uh", "um", "cl")) +
  labs(x = "PINTs", y = "Count") +
  facet_wrap(vars(speaker), ncol = 5, labeller = labeller(speaker = sp.lab)) +
  theme(panel.spacing.x = unit(1, "lines"))
```

## The following plots are saved into the output directory and used for the poster publication
### Between-/within- speaker barplot
```{r, warning=FALSE}
plot_between_lectures <- ggplot(df_yale_pints, aes(x = segment, fill = session)) + 
  geom_bar(position = position_dodge()) +
  scale_x_discrete(limits = c("sil", "in", "ex", "uh", "um", "cl"), expand = c(0, 1)) +
  scale_fill_discrete(name = "Session", labels = c("First", "Second", "Third")) +
  labs(x = "PINTs", y = "Count") +
  facet_wrap(vars(speaker), ncol = 5, labeller = labeller(speaker = sp.lab)) +
  theme(panel.spacing.x = unit(1, "lines"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent"),
        plot.background = element_rect(fill = "transparent", color = NA),
        axis.text.x = element_text(angle = 45),
        strip.background = element_blank(),
        text = element_text(size = 32))

ggsave(here::here("output/plot_between_lectures.png"), plot_between_lectures, width = 18, height = 10, units = c("in"))
```

### PINTs proportion barplot
```{r, warning=FALSE}
plot_prop_bar <- ggplot(df_yale_prop, aes(x = segment, y = prop_total)) +
  geom_bar(stat = "identity",
           fill = "#F8766D", 
           color = "black") + 
  geom_text(aes(label = round(prop_total, 1)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5,
            size = 14) +
  ylim(0, 15) + 
  labs(x = "PINTs", y = "% of Total Speaking Time") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent"),
        plot.background = element_rect(fill = "transparent", color = NA),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.text = element_text(size = 28),
        axis.title = element_text(size = 36),
        axis.title.x = element_blank())

ggsave(here::here("output/plot_prop_bar.png"), plot_prop_bar, width = 18, height = 10, units = c("in"), bg = "transparent")
```

### PINTs pie chart
```{r, warning=FALSE}
df_pie_chart <- data_frame(
  Label = c("PINTs", "Speech"),
  value = c("29.4%", "70.6%")
)

plot_pie_chart <- ggplot(df_pie_chart, aes(x = "", y = value, fill = Label)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(aes(label = value), 
            position = position_stack(vjust = 0.5),
            size = 14) +
  coord_polar("y", start = 360) +
  theme_void() +
  theme(legend.key.size = unit(.75, "in"),
        legend.title = element_text(size = 36),
        legend.text = element_text(size = 28))

ggsave(here::here("output/plot_pie_chart.png"), plot_pie_chart, width = 18, height = 10, units = c("in"))
```

### PINTs duration boxplot
```{r, warning=FALSE}
df_yale_boxplot <- df_yale_pints
df_yale_boxplot$segment <- factor(df_yale_boxplot$segment, levels = c("sil", "in", "uh", "ex", "um", "other", "cl"))

plot_boxplot <- ggplot(df_yale_boxplot, aes(x = segment, y = dur, fill = "#F8766D")) + 
  geom_boxplot() +
  ylim(0,1.0) +
  xlab("PINTs") +
  ylab("Duration (s)") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent"),
        plot.background = element_rect(fill = "transparent", color = NA),
        legend.position = "none",
        axis.text = element_text(size = 28),
        axis.title = element_text(size = 36),
        axis.title.x = element_blank())

ggsave(here::here("output/plot_boxplot.png"), plot_boxplot, width = 18, height = 10, units = c("in"))
```


